---
title: "EPI packages for R"
---

[epiverse](https://economic.github.io/epiverse/)

-   now you can load all of EPI's R packages in one go! `epiverse` includes the libraries you know and love (realtalk, epidatatools, and epiextractr), as well as two hot new packages!
    -   [prefab](https://economic.github.io/prefab/) is an opinionated theme system for R project scaffolding

    -   [swadlr](https://economic.github.io/swadlr/) allows users to pull and manipulate data from the [State of Working America Data Library](https://data.epi.org/)

[realtalk](https://economic.github.io/realtalk/)

-   realtalk makes it easy to load and use common US price indexes in R.

[epidatatools](https://economic.github.io/epidatatools/)

-   epidatatools contains functions we find useful at [EPI](https://epi.org/) that don't have exact analogues elsewhere in the R package ecosystem.

[epiextractr](https://economic.github.io/epiextractr)

-   use epiextractr to download CPS data from the [EPI microdata site](https://microdata.epi.org/).

### To install:

You can install any of these packages from the EPI R-universe [repository](https://economic.r-universe.dev/). For example, the `epiverse` package can be installed with:

```{r epiverse, eval=FALSE, message=FALSE}
install.packages("epiverse", repos = c("https://economic.r-universe.dev",) getOption("repos")))
```

Similar code is available on each package's GitHub page.

However, for easier syntax and updates, consider adding the repository to the "repos" option in your `.Rprofile` file:

```{r renviron, eval=FALSE, message=FALSE}
# Example .Rprofile  
options(repos = c(
    epi = "https://economic.r-universe.dev",
    getOption("repos")
))
```

After making those edits and restarting R, you can just use `install.packages("epiverse")` to update packages to the latest version, or install new EPI packages.

To find and edit your `.Rprofile`, use `usethis::edit_r_profile()`.