---
title: "Load EPI CPS Extracts via epiextractr"
---

**Note:** Users will need to install `epiextractr` for this example. Refer to [EPI packages for R](epi_libraries.qmd) for installation instructions.

```{r Internal load libraries, include=FALSE, message=FALSE}
library(tidyverse)
library(epiextractr)
library(here)
```

```{r download cps, eval=FALSE, include=FALSE, message=FALSE}

# download CPS ORG files  
download_cps(sample = 'org', extracts_dir = here('data/cps'))

# Users can also download CPS Basic (4085.3 MB) and CPS May (38.8 MB) files 
download_cps(sample='basic', extracts_dir = here('data/cps'))

download_cps(sample='may', extracts_dir = here('data/cps'))

```

### Load required libraries

```{r Load libraries, eval=FALSE, message=FALSE}
library(tidyverse)
library(epiextractr)
```

### Download CPS files

```{r Download CPS files from microdata site, eval=FALSE, message=FALSE}

# download CPS ORG files  
download_cps( sample = 'org', extracts_dir = 'C:/YOUR_PATH/cps', overwrite = TRUE)

# download CPS Basic (4085.3 MB) 
download_cps(sample='basic', extracts_dir ='C:/YOUR_PATH/cps', overwrite = TRUE)

# download CPS May (38.8 MB)
download_cps(sample='may', extracts_dir = 'C:/YOUR_PATH/cps', overwrite = TRUE)

```

This will download the latest EPI CPS ORG extracts in .feather format from [https://microdata.epi.org](https://microdata.epi.org/) and place them in the directory `C:\data\cps`.

### Load your CPS extracts!

After the data is downloaded, load a selection of CPS data for your analysis:

```{r Load ORG, eval=FALSE, message=FALSE}

org <- load_cps("org", 2010:2022, year, orgwgt, wage, age, statefips, wbho, 
                .extracts_dir = 'C:/YOUR_PATH/cps')

```

### (Optional) Set and forget CPS data files by creating/editing an .Renviron file

To simplify usage, you can omit the `.extracts_dir` argument by setting the environment variables to your extracts directory. This allows you to call CPS extracts from a single, dedicated folder, eliminating redundant downloads of CPS files.

Edit your .renviron file easily by installing the [usethis](https://usethis.r-lib.org/) package `install.packages("usethis")`

```{r Access your .Renviron file, eval=FALSE}
# Load use this
library(usethis)

# run the edit_r_environ() command from the usethis package
edit_r_environ()

# Paste the file paths that point to your CPS files into the .renviron file in the following format:

EPIEXTRACTS_CPSBASIC_DIR = 'C:/YOUR_PATH/cps'
EPIEXTRACTS_CPSMAY_DIR = 'C:/YOUR_PATH/cps'
EPIEXTRACTS_CPSORG_DIR = 'C:/YOUR_PATH/cps'

```

After editing your `.Renviron` file, save your changes and restart R to apply them.

If you've set your .Renviron file paths to point to the folder containing your CPS files, you can omit the `.extracts_dir()` command

```{r eval=FALSE, message=FALSE}

org <- load_cps("org", 2010:2022, year, orgwgt, wage, age, statefips, wbho)
```