---
title: "R Installation Webinar"
format: html
author: "Daniel Perez & Emma Cohn"
---

## Welcome to the R Installation Webinar landing page!

This training and code workflow was originally delivered as an EARN Talk on July 29th, 2025.

Missed this talk? See the recording here: [Preparing for Analysis with R: A guided tutorial for installing R and RStudio](https://us06web.zoom.us/rec/share/NfQbs0nb4BWsxZQw8NoKzJV3fhowtRXxKwc_UqjYFYgCRkgmywnCw4uw9lESnBL7.Xk7du8aT4P-K7YLB).

**Passcode:**

[Presentation slides](https://github.com/Economic/earn_code_library/blob/main/supp_files/R_prep_webinar_slides.pdf)

## Workflow example

This example demonstrates a simple data analysis workflow.

In short, we will use the Tidyverse package to load a dataset from a .csv file, describe and analyze the dataset, and export a final analysis to a new .csv file.

## Load a library

Recall, you can install R packages using the `install.packages()` command.

```{r, eval=FALSE}
install.packages('tidyverse')
```

After installing a package, you can load it using the `library()` command.

```{r Load libraries, message=FALSE}
# Loading a library 
library(tidyverse)
```

## Read data from a csv file

1.  Download the dataset at this link: [counties_per_capita_income.csv](https://github.com/Economic/earn_code_library/blob/main/data/counties_per_capita_income.csv)

2.  Place the dataset in your root directory In this case, our data lives in a folder named "data" within this root directory. You can name your folder whatever you'd like!

3.  Use the `read.csv()` function to load the data into R.

```{r read CSV}
counties_income <- read.csv("data/counties_per_capita_income.csv")
```

## Descriptive analysis

Some useful commands to describe your dataset

```{r descriptive commands}
# Print the number of rows and columns in your table 
dim(counties_income)

# View the top 5 rows of your dataset
head(counties_income)

# List the variable names of your dataset
names(counties_income)

# View a transposed table of your data
glimpse(counties_income)

# Generate summaries for each of your variables
summary(counties_income)

# Generate a summary of one variable in your datset using the $ operator
summary(counties_income$household_income)

```

## Simple summary statistics

Now that we've loaded our dataset, let's run a simple analysis.

The following code block uses the `summarize()` function to "collapse" variables from the `counties_income` dataset into a smaller table with summary statistics. Specifically, we calculate the median household income and a count of counties for each state.

In this example

-   `median(household_income)` calculates the median income.

-   `n()` is a function that counts the number of observations in our sample.

```{r Summary stats}

summarize(counties_income,
            med_hhinc = median(household_income),
            n=n())

```

We can also create summary statistics by specific groupings. For example, if we want to calculate the median household income for each state, we can use the `.by=` argument inside `summarize()`

```{r groupings}

summarize(counties_income, 
          med_hhinc = median(household_income), 
          n = n(), 
          .by = states)
  
```

This tells R to calculate the summary statistics separately for each unique value in the `states` column

## Creating new variables with mutate

The `mutate()` function adds new variables to a dataset without collapsing it, unlike `summarize()` which reduces the dataset to summary rows.

In the example below, we use `mutate()` to create a new column called `rank`, which ranks counties from highest to lowest based on their per capita income (pci).

-   rank(-pci) ranks the values of pci in descending order (the minus sign indicates descending).

-   The result is a new dataset `ranked_income` where each county keeps its original data, now with an added rank column.

```{r Summarize by groups}

ranked_income <- mutate(counties_income, rank = rank(-pci))

# Print the top 5 rows of our new dataframe/table
head(ranked_income)

```

## How to look up a function's arguments

Not sure what a function does or what arguments it accepts? You can use the help command in R to look it up!

Type a question mark (?) followed by the function name in your console or code chunk:

```{r seeking help}

?mutate

?summarize

# alternatively, you can type
help(summarize)

```

This opens the help file for the function, which includes a short description of what the function does, a list of arguments you can use, and examples showing how to apply it. This is a great habit to adopt when trying new functions!

## Using pipes to create multi-line commands

In R, the pipe operator `|>` allows you to chain together multiple steps of a data transformation in a readable, top-to-bottom format. Each step passes its result to the next command. This makes your code easier to read and avoids creating lots of intermediate objects.

Here's an example using the \|\> pipe to filter and summarize data across multiple lines:

```{r Multi-line commands}

# Executing a multi-line command using pipes 
cleaned_data <- counties_income |>
    # Here we use the filter() command to restrict our sample to a few states
    filter(states %in% c('New York', 'California', 'North Carolina' )) |>
    # Use summarise (Note, summarize() and summarise() are interchangeable. British spelling fans, rejoice!)
    summarise(mean_hhinc = mean(household_income), 
              n = n(), 
              .by = states)

```

What this code does:

1.  Starts with the `counties_income` dataset

2.  Filters it to only include counties in New York, California, and North Carolina

3.  Summarizes the data by state, calculating:

    1.  The mean household income

    2.  The number of counties

## Exporting your results

Once you've created a cleaned or summarized dataset, you may want to save it to a file. The `write.csv()` function lets you export your data to a `.csv` file that you can open in Excel or share with others.

```{r Write to csv}

# Writing a csv to a path
write.csv(cleaned_data, "output/state_incomes.csv")
```

This line:

-   Writes the `cleaned_data` dataset to a file called `state_incomes.csv`

-   Saves it in the `output/` folder within your project directory

**Note**: Make sure the `"output"` folder already exists in your project. R will return an error if the folder doesn’t exist.

## Additional resources

As you embark on your R journey, check out the [Other Resources](resources.qmd) page for additional materials!

Happy coding!