---
title: "Coding race and ethnicity variables"
author: "EPI data team"
---

```{r, load library, eval=FALSE, include=FALSE, message=FALSE}
library(tidyverse)
```

## Recoding for ASEC

```{r, asec race coding, eval=FALSE, include=FALSE, message=FALSE}
mutate(
    wbhao = case_when(
        hispan >= 100 & hispan <= 612 ~ 3, # Hispanic or latino
        race == 100 ~ 1, # White
        race %in% c(200, 801, 805, 806, 807, 810, 811, 814, 816, 818) ~ 2, # Black
        race %in% c(650, 651, 652, 803, 804, 808, 809, 812, 813, 817, 819) ~ 4, # AAPI
        TRUE ~ 5
    ),
    wbhao = haven::labelled(wbhao, c(
        "White" = 1, 
        "Black" = 2, 
        "Hispanic" = 3, 
        "AAPI" = 4, 
        "Other" = 5
    ))
)
```

## Recoding for ACS (via IPUMS)

There are two ways to recode ACS race categories, the first is using the `race` and `raced` variables (`raced` gets automatically downloaded when you include the race variable in your sample), and the second is using `race` and the 'rac' variable series (i.e. `racblk`, `racasian`, `racpacis`, `racamind`).

Using `race` and `raced`:

```{r, acs race coding 1, eval=FALSE, include=FALSE, message=FALSE}
mutate(
    wbhao = case_when(
        hispan %in% c(1,2,3,4) ~ 3, 
        race == 1 ~ 1, 
        race == 2 ~ 2,
        raced >= 830 & raced <= 845 ~ 2,
        raced >= 901 & raced <= 904 ~ 2,
        raced >= 930 & raced <= 936 ~ 2,
        raced >= 950 & raced <= 955 ~ 2,
        raced >= 970 & raced <= 973 ~ 2,
        raced >= 980 & raced <= 983 ~ 2,
        raced %in% c(917, 985, 986, 990, 991) ~ 2,
        race %in% c(4,5,6) ~ 4,
        raced >= 810 & raced <= 825 ~ 4,
        raced >= 850 & raced <= 855 ~ 4,
        raced >= 860 & raced <= 899 ~ 4,
        raced >= 910 & raced <= 915 ~ 4,
        raced >= 920 & raced <= 927 ~ 4,
        raced >= 940 & raced <= 944 ~ 4,
        raced >= 960 & raced <= 964 ~ 4, 
        raced %in% c(905, 974, 975, 976, 984) ~ 4,
        TRUE ~ 5
    ),
    wbhao = labelled(wbhao, c(
        "White" = 1, 
        "Black" = 2, 
        "Hispanic" = 3, 
        "AAPI" = 4, 
        "Other" = 5
    ))
)
```

Using `race` and the 'rac' variable series

```{r, acs race coding 2, eval=FALSE, include=FALSE, message=FALSE}
mutate(
    wbhao = case_when(
        hispan %in% c(1,2,3,4) ~ 3, 
        race == 1 ~ 1, 
        racblk == 2 ~ 2,
        racasian == 2 ~ 4,
        racpacis == 2 ~ 4,
        TRUE ~ 5
    ),
    wbhao = labelled(wbhao, c(
        "White" = 1, 
        "Black" = 2, 
        "Hispanic" = 3, 
        "AAPI" = 4, 
        "Other" = 5
    ))
)
```