---
title: "Coding employment outcomes"
author: "EPI data team"
---

**Note:** Users will need to install `epiextractr` for this example. Refer to [EPI packages for R](epi_libraries.qmd) for installation instructions.

```{r Internal load libraries, include=TRUE, message=FALSE}
library(tidyverse)
library(epiextractr)
```

## Employment-to-population ratios (EPOPs)

This code will allow the user to reproduce [employment-to-population ratios](https://data.epi.org/labor_force/labor_force_emp/line/year/national/percent_emp/overall?timeStart=1976-01-01&timeEnd=2025-01-01&dateString=2025-01-01&highlightedLines=overall). Please note that data with missing observations will return missing, therefore we must handle missing data directly. It is common practice to use \`filter()’ to remove missing data directly with the most control. 

```{r calculate epop, eval=FALSE, include=TRUE, message=FALSE}
load_basic(2020:2024, year, age, emp, basicwgt) |>  
  filter(age >= 16, !is.na(emp))  |>  
  #note: the denominator is based on number of months that are included in the sample
  #note: since we are getting annual estimates, divide the weight by 12
  mutate(adj_wgt = basicwgt/12) |>  
  summarise( 
    epop = weighted.mean(emp, w = adj_wgt), 
    .by = year) 
```

## Labor force participation rates

This code will allow the user to reproduce the [labor force participation rates](https://data.epi.org/labor_force/labor_force_lf/line/year/national/percent_lf/overall?timeStart=1976-01-01&timeEnd=2025-01-01&dateString=2025-01-01&highlightedLines=overall). Similar to EPOPs, we must be careful how we handle missing data, however unlike EPOPs this code uses the `na.rm = TRUE` argument inside the operation. Removing missing data this way gives the user less control over which observations are dropped.

```{r calculate lfpr, eval=FALSE, include=TRUE, message=FALSE}
load_basic(2020:2024, year, age, lfstat, basicwgt) |>  
  filter(age >= 16)  |>  
  mutate( 
    #note: the denominator is based on number of months that are included in the sample
    #note: since we are getting annual estimates, divide the weight by 12
    adj_wgt = basicwgt/12, 
    lfp = if_else(lfstat == 1 | lfstat == 2, 1, 0)) |>  
  summarise( 
    lfpr = weighted.mean(lfp, w = adj_wgt, na.rm = TRUE), 
    .by = year) 
```

## Unemployment rate

This code will the user to reproduce the [unemployment rates](https://data.epi.org/labor_force/labor_force_unemp/line/year/national/percent_unemp/overall?timeStart=1976-01-01&timeEnd=2025-01-01&dateString=2025-01-01&highlightedLines=overall). 

```{r calculate urate, eval=FALSE, include=TRUE, message=FALSE}
load_basic(2020:2024, year, age, unemp, lfstat, basicwgt) |>  
  filter(age >= 16, lfstat != 3)  |>  
  #note: the denominator is based on number of months that are included in the sample
  #note: since we are getting annual estimates, divide the weight by 12
  mutate(adj_wgt = basicwgt/12) |>  
  summarise( 
    urate = weighted.mean(unemp, w = adj_wgt), 
    .by = year) 
```